% Sample University of Calgary Thesis
% This file contains the FRONT MATTER other thna the title page

\chapter{Abstract}
The divisor class group of a hyperelliptic curve over a field is a finite
abelian group at the center of many open questions in algebraic
geometry and number theory. Sutherland~\cite{Sutherland_g3_2019} surveys some of
these, including the computation of the associated $L$-functions and zeta
functions used in his investigation of Sato-Tate
distributions~\cite{Sutherland_sato_2016}. Many of these problems lend
themselves to empirical investigation, and as emphasized by Sutherland, fast
arithmetic in the divisor class group is crucial for their efficiency. Indeed,
implementations of these fundamental operations are at the core of the algebraic
geometry packages of widely-used computer algebra systems such as Magma and
Sage. 

This thesis provides contributions to improve the efficiency of divisor class
arithmetic on hyperelliptic curves, with special attention to the often ignored
split model cases and to genus 2 and 3. There are two main contributions: the
introduction of ``Balanced NUCOMP'' for improved arithmetic on curves given by
split models of arbitrary genus, and improved explicit formulas for genus 2
(ramified and split models) and genus 3 (split models) based on Balanced NUCOMP
for split models and NUCOMP for ramified models. Empirical analysis, using a
complete Magma implementation and testing suite, is conducted with all
contributed algorithms to provide proof of correctness and comparisons to
previous best. Our results show that Balanced NUCOMP does offer improvements to
split model arithmetic, further narrowing the performance gap between split and
ramified models. Our explicit formulas require fewer field operations, most
notably significantly fewer additions, than previous best for computing
arithmetic on genus 2 (ramified and split models), and genus 3 (split models).
Our empirical results demonstrate that our formulas yield the fastest
general-purpose arithmetic for these cases.

\chapter{Preface}

This thesis is original, unpublished, independent work by the author,
Sebastian, A, Lindner.


%Portions of the introductory text of Chapter~1 are used with
%permission from Smith et al. (2015) of which I am an author. Table~1.1
%is modified from Supplementary Table~3 in Smith et
%al. (2014). Chapter~2 of this thesis has been published as J.~Smith
%and J.~Doe, ``Title of Article''. Journal Name, vol.~1, issue~1.


  
\chapter{Acknowledgments}

I would like to thank my supervisor, Dr. Michael J. Jacobson, for his consistent support
and guidance throughout all stages of my academic growth and this dissertation.
Furthermore I would like to acknowledge and thank Dr. Renate Scheidler, Dr. Laurent Imbert, 
Randy Yee, and Parthasarathi Das for their support and for meaningful
conversations that contributed to my academic growth and the completion of this
work. Finally, thank you to my wife for her seemingly endless patience, support
and encouragement. 

\tableofcontents

% If you have no tables, delete the next line

\listoftables

% If you have no figures, delete the next line
\addcontentsline{toc}{chapter}{List of Algorithms}
\listofalgorithms

% Consult Ch 9 of the memoir class manual on how to set up other
% content lists. Note that memoir does not automatically clear the
% page for these. ucalgmthesis fixes this for the default table of
% contents and lists of tables and figures, but not for anything you
% define

%\listoffigures